---
title: "p8105_hw2_khs2318"
author: "Kayla Schiffer-Kane"
date: "2023-10-04"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

```{r libraries}
library(tidyverse)
library(readxl)
```

# Problem 1 

Dataset is FiveThirtyEight data 

## Clean pols_month

Clean the data in pols-month.csv. Use `separate()` to break up the variable mon into integer variables `year`, `month`, and `day`; replace `month` number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the `day` variable.
 

```{r clean_pols_month}
pols_month =
  read_csv('./data/fivethirtyeight_datasets/pols-month.csv') |>
  separate(mon, c("year","month","day")) |>
  mutate(month = month.name[as.numeric(month)]) |>
  mutate(president = recode(prez_dem, "1" = "dem", "0" = "gop")) |>
  select(-prez_dem,-prez_gop, - day)
  
head(pols_month)
```


## Clean snp 

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r clean_snp}
snp =
  read_csv('./data/fivethirtyeight_datasets/snp.csv') |>
  separate(date, c("month","day","year"), sep = '/') |>
  mutate(month = month.name[as.numeric(month)]) |>
  mutate(year = if_else(year > 15, paste0("19",year), paste0("20",year))) |> 
  #arrange(year, month) |>
  select(year,month,close)
tail(snp)
```


## Clean Unemployment
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.


```{r clean_unemployment}
unemployment =
  read_csv('./data/fivethirtyeight_datasets/unemployment.csv') |> 
  pivot_longer(Jan:Dec,
               names_to = "month",
               values_to = "percentage") |>
  janitor::clean_names() |>
  mutate(month = month.name[match(month, month.abb)]) |> # Jan -> January 
  mutate(year = as.character(year))
unemployment
```

## Merge all data
Required casting year in unemployment to character

```{r merge_538}
data = 
  left_join(pols_month, snp) |>
  left_join(x = _, y = unemployment)
```

The resulting dataset has `r nrow(data)` rows and `r ncol(data)` columns. 

* Pols_month has  `r nrow(pols_month)` rows, and we created `president` which tells us if the president was a gop/dem in a given year. This column was created from a combination of `prez_gop` and `prez_dem`, which I tested to make sure are truly mutually exclusive. Years include `r min(pols_month$year)` to `r max(pols_month$year)`
* SNP has `r nrow(snp)`, with data from `r min(snp$year)` to `r max(snp$year)`. Average close price for these years is `r filter(snp, close >= 0) |> pull(close) |> mean()`
* Unemployment has unemployment rates from `r min(unemployment$year)` to `r max(unemployment$year)`. Average unemployment for these years is `r filter(unemployment, percentage >= 0) |> pull(percentage) |> mean()`
* Because the years don't overlap exactly, we'll miss some data for certain years - hence the "NAs" in the merged dataset - 36 values are missing SNP close price `close`, and 12 are missing unemployment rates `percentage`.


# Problem 2 

## Load and Clean Data
Read and clean the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data

* The last row contains sums of the previous so should be excluded. This is the only row where date is missing, so excluding it by filtering out where Month is null. 

```{r load_trashwheel}
trashwheel_data = read_excel("./data/202309 Trash Wheel Collection Data.xlsx",
                             sheet = "Mr. Trash Wheel",
                             skip = 1,
                             range = cell_cols("A:N")) |>
  filter(!is.na(Month)) |> 
  janitor::clean_names()
skimr::skim(trashwheel_data)

trashwheel_data
```

## Calculate homes_powered

The data include a column for the (approximate) number of homes powered. This calculation is described in the `Homes powered note`, but not applied to every row in the dataset. Update the data to include a new `homes_powered` variable based on this calculation.

* From excel: Homes Powered - Each ton of trash equates to on average 500 kilowatts of electricity. An average household will use 30 kilowatts per day.

```{r homes_powered}
kw_per_ton = 500
kw_per_house = 30
trashwheel_data = trashwheel_data |>
  rename(homes_powered_orig = homes_powered) |>
  mutate(homes_powered = (weight_tons * kw_per_ton / kw_per_house))
```

*Checking* to see if the calculation looks similar to the original `homes` column. Using only where the homes_powered from original dataset was  calculated, the average number of homes powered is `r trashwheel_data |> filter(homes_powered_orig > 0) |> pull(homes_powered_orig) |> mean() |> round(2)` compared to our calculation, `r pull(trashwheel_data, homes_powered) |> mean() |> round(2)`. These are fairly equivalent. 

## Remaining
Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?



